Boolean Operations for Surfaces in VTK Without External Libraries

### 2011-05-10 Document/LaTeX/Article.tex
Added figures.

While it seems possible to use the class \code{vtkClipPolyData} together with \code{vtkImplicitPolyData} to extract the necessary portions of each input surface to produce the boolean surfaces, there is a problem. \code{vtkClipPolyData} treats the implicit function as piecewise linear within each cell of the \code{vtkPolyData} being clipped. In general, implicit functions are not guaranteed to be piecewise linear in each cell, and the same is true for the signed distance field. The result is that clipped surfaces produced by \code{vtkClipPolyData} will not be clipped at the true zero level of the distance field. Consequently, the clipped portions of each input geometry will not match at the zero level where they should fuse seemlessly [INSERT FIGURE].

@@

Identifying the surface intersection between two surface meshes is the most complex part of the boolean operations algorithm. It involves identifying the intersection between each triangle in the first mesh with all triangles in the second mesh that intersect it. A triangle-triangle intersection test yields two endpoints (possibly the same) of the actual line segment defining the intersection when the triangles intersect. The intersection between overlapping and co-planar triangles is more complicated; \code{vtkPolyDataIntersection} does not explicitly handle intersections between two coplanar triangles.

@@

The entries in the \code{std::multimap} described above are used to split the fully connected intersection lines to respect the topology of each input mesh. Because the end points of the intersection lines are merged during construction, they need to be split at locations where the mesh is split. For example, a mesh representing a cube often has duplicate points at the corners to store three different normals, one for each face incident to the point. These normals enable the appearance of sharp edges on the mesh. If the surface intersection on a cube splits an edge that connects two of these corners, then two points should be inserted at the split point rather than one to ensure that two face normals are defined at the split point.

@@

Splitting proceeds on a cell-by-cell basis. For each cell that needs splitting, the following data is gathered:

\begin{enumerate}
\item The three points that define the cell.
\item The boundary lines of the cell.
\item The split intersection lines that lie in the cell (if any).
\item The points that define the lines in 3.
\item Points from neighboring cells that lie on a cell edge but which were not already added in 4.
\end{enumerate}

>>>>

While it seems possible to use the class \code{vtkClipPolyData} together with \code{vtkImplicitPolyData} to extract the necessary portions of each input surface to produce the boolean surfaces, there is a problem. \code{vtkClipPolyData} treats the implicit function as piecewise linear within each cell of the \code{vtkPolyData} being clipped. In general, implicit functions are not guaranteed to be piecewise linear in each cell, and the same is true for the signed distance field. The result is that clipped surfaces produced by \code{vtkClipPolyData} will not be clipped at the true zero level of the distance field. Consequently, the clipped portions of each input geometry will not match at the zero level where they should fuse seamlessly. XXXX Figure

@@

Identifying the surface intersection between two surface meshes is the most complex part of the boolean operations algorithm. It involves identifying the intersection between each triangle in the first mesh with all triangles in the second mesh that intersect it. A triangle-triangle intersection test yields two endpoints (possibly the same) of the actual line segment defining the intersection when the triangles intersect. If the endpoints are the same, then the intersection line is degenerate, and it is not added to the set of intersection lines. The intersection between overlapping and co-planar triangles is more complicated; \code{vtkPolyDataIntersection} does not explicitly handle intersections between two coplanar triangles.

@@

The entries in the \code{std::multimap} described above are used to split the fully connected intersection lines to respect the topology of each input mesh. Because the end points of the intersection lines are merged during construction, they need to be split at locations where the mesh is split. For example, as shown in Figure \ref{fig:BoxVectors}, a mesh representing a cube often has duplicate points at the corners to store three different normals, one for each face incident to the point. These normals enable the appearance of sharp edges on the mesh. If the surface intersection on a cube splits an edge that connects two of these corners, then two points should be inserted at the split point rather than one to ensure that two face normals are defined at the split point.

\begin{figure}
\centering
\includegraphics[scale=0.25]{Figures/BoxNormals}
\itkcaption{Splitting a mesh with duplicate normals. The box object has three duplicate points at each corner each of which has a different normal (gray arrows). The surface intersection between the box in the front (gray, solid) and the box in the back (white, outline) is shown in green. Points from the intersection lines that lie on the duplicate edges of the box are also duplicated by the mesh splitting procedure in \code{vtkPolyDataIntersection}. The normal at a duplicated point, along with other point data, is interpolated from the cell that contains the intersection line that references the point.}
\label{fig:BoxNormals}
\end{figure}

@@

Splitting proceeds on a cell-by-cell basis. For each cell that needs splitting, the following points and lines are gathered:

\begin{enumerate}
\item The three points that define the cell (green spheres in Figure \ref{fig:CellSplitSetup}).
\item The boundary lines of the cell (red lines in Figure \ref{fig:CellSplitSetup}).
\item The split intersection lines that lie in the cell (if any) (white lines in Figure \ref{fig:CellSplitSetup}).
\item The points that define the lines in 3 (purple spheres in Figure \ref{fig:CellSplitSetup}).
\item Points from neighboring cells that lie on a cell edge but which were not already added in 4 (orange sphere in Figure \ref{fig:CellSplitSetup}).
\end{enumerate}

\begin{figure}
\centering
\includegraphics[scale=0.25]{Figures/CellSplitSetup}
\caption{The points and lines required for splitting a cell.}
\label{fig:CellSplitSetup}
\end{figure}
